Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-02-10 ThesisJP/Background/background.tex
Continued writing background information.

In this chapter the background information needed in order to understand the topics of this thesis and its papers is given. First, in Section \ref{sec:gpgpu}, general-purpose programming on graphics processing units (\nom{GPGPU}{General-purpose programming on graphics processing units}) is reviewed, followed by an introduction to medical ultrasound imaging in Section \ref{sec:ultrasound}. Finally, the concepts of volume rendering and field simulations are presented in Section \ref{sec:volren} and \ref{sec:field}.

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spins around the fact that chip manufactures hit a frequency-wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single core, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch" era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation where also growing at the same rate, until the level of cooling required finally became to much in 2004.  From that point, chip manufactures had to pursue a different approach  than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU chip. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to continue to gain performance. It was a concurrency revolution, and  the free lunch was over.

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{img/free_lunsh.png}
\caption{Illustration of several transitions currently taking place in the computer industry. \todo{Site Herb Sutter - Welcom to the jungle http://herbsutter.com/welcome-to-the-jungle/}}
\label{fig:jungle}
\end{figure}

Around the same time as CPU manufactures hit the frequency wall people started experimented with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had previously been a fixed function pipeline, where e.g. vertices and texture coordinates where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that hidden in the GPU was processing power an order of magnitude larger than what the CPU could provide at the same time. Actually, at this time, the GPU had already gone multi-core, driven by the increasing demand for realistic computer games. Multiple cores was somewhat easier to achieve with the GPU since the rendering problem is inherently parallel. The pixels in the rendered image can be calculated in any order, and one pixel is therefore an concealed unit which the GPU scheduler is free to issue at any time to whatever core that is available. It it is also much less costly to switch between a pixel-thread for the GPU then a program thread for the CPU. Since there was only one program running on the GPU, the fixed function pipeline, GPU designers had a good idea of how the program would execute, and could therefore skip advance feature as predicting execution paths and different levels of data caching.  The released silicon was used to make more cores.

It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.

@@

Early gpgpu programming required expert knowledge about graphics and the remaining fixed funtions in the pipeline. New programming languages where needed. OpenCL and CUDA. C++ AMP.

@@

Compare CPU and GPU ALUs.

Add sentence about siemens and supersonic imagine scanners. They use GPUs? What does this mean? FPGAs and ASICs.

@@

Small section about heart anatomy.

@@

>>>>

In this chapter the background information needed in order to understand the topics of this thesis and its papers is given. First, in Section \ref{sec:gpgpu}, general-purpose programming on graphics processing units (\nom{GPGPU}{General-purpose programming on graphics processing units}) is reviewed, followed by an introduction to medical ultrasound imaging in Section \ref{sec:ultrasound}. Finally, the concepts of volume rendering and ultrasound field simulations are presented in Section \ref{sec:volren} and \ref{sec:field}.

\section{General-purpose computing on graphics processing units}\label{sec:gpgpu}
In a recent essay, Herb Sutter, a leading authority on software development, reviews the current state of the software and computer industry \cite{HerbSutter}. The essay bears the name "Welcome to the Jungle" and is a sequel to the essay "The Free Lunch is Over" from 2004 \cite{HerbSuttera}. The two essays spins around the fact that chip manufactures hit a frequency-wall in the beginning of the 21st century. Until then, all mainstream computer programs where typically running in a single thread on a single core, and a programmer could expect the software to annually gain performance without touching the code. This was the "free-lunch"-era as depicted in Fig.\,\ref{fig:jungle}. The increase in processing power came mostly from an ever-growing clock frequency. However, power consumption and heat generation where also growing at the same rate, until the level of cooling required finally became to much in 2004.  From that point, chip manufactures had to pursue a different direction  than just increasing the clock frequency. The immediate solution was to embed several cores in one CPU chip. This insured continuation of Moore's law, since the number of transistors per chip could continue to grow. For programmers, multi-core CPUs meant that computer programs now had to be multi-threaded in order to continue to gain performance. It was a concurrency revolution, as Herb Sutter named it, and  the "free lunch" was over.

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{img/free_lunsh.png}
\caption{Illustration of transitions between different eras in the computer industry. Currently there are three ongoing transitions, multi-core, heterogeneous-core and cloud-core. Image courtesy of Herb Sutter (herbsutter.com).}
\label{fig:jungle}
\end{figure}

Around the same time as CPU manufactures hit the frequency wall, people started experimenting with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and textures where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that hidden in the GPU was processing powers an order of magnitude larger than what the CPU could provide at the same time (See Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at this time the GPU had already gone multi-core, driven by an ever-increasing demand for more realistic computer games. What also made this possible was the inherent parallel nature of the rendering problem. When geometry is projected and rasterized the same instruction is typically needed for a lot of data at the same time. This makes a special kind of instruction, Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction, especially suited for this kind of problem. The rendering pipeline also exhibits minimal execution path branching, and data are typically used once. Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching. This reduced the silicon footprint of each core and made it possible to add multiple SIMD cores to the same chip before CPU designers where able to do the same. Today, even if more advanced caching has added to GPUs, this is still the main reasons why GPUs have higher peak performance than CPUs (as shown in Fig.\,\ref{fig:cpu_vs_gpu}).

The first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU), wanted to use the GPU to accelerate computationally demanding algorithms.

FPGAs (Field-programmable gate array), APU (Accelerated processing unit).

@@

\subsection{Comparing CPU and GPU performance} 
It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.

It it is also much less costly to switch between a pixel-thread for the GPU then a program thread for the CPU.

Compare CPU and GPU ALUs. SSE, AVX ... The GPU has super-wide registers. The SM in Fermi have 32 cores or 1024 bit registers. The SMX in Kepler have 192 cores. 6144 bit registers. SSE (Streaming SIMD Extensions) instructions have 128 bit registers. AVX (Advanced Vector Extensions) have 256 bit registers. AVX-512 has 512 bit registers. Single instruction, multiple data (SIMD).

\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{img/cpu-vs-gpu.png}
\caption{Development in theoretical peak throughput, single precision (SP) and double precision (DP), the last decade for Intel CPUs and Nvidia GPUs. Image courtesy of Michael Galloy (michaelgalloy.com).}
\label{fig:cpu_vs_gpu}
\end{figure}

\subsection{Programming the GPU}
Early gpgpu programming required expert knowledge about graphics and the remaining fixed funtions in the pipeline. New programming languages where needed. OpenCL and CUDA. C++ AMP.

@@

Hopefully, in the future, GPU programming will be the job of e.g. c++ compiler, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. (In the same way the compiler is able to translate c++ code into SSE and AVX instructions.)

\subsection{Accelerators in medical ultrasound imaging }
Add sentence about siemens and supersonic imagine scanners. They use GPUs? What does this mean? FPGAs and ASICs.

@@

@@

\subsection{Cardiac anatomy}
Small section about heart anatomy.

### 2014-02-10 ThesisJP/Introduction/introduction.tex
Continued writing background information.

\subsection{Paper II}
\textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
\textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C\:Nilsen, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. \todo{add pages}, Jan. 2014.}\\\\
The second paper aims at implementing real-time Capon beamforming for cardiac ultrasound imaging. Achieving this will facilitate further study of the method's \textit{in vivo} performance. This is something which has been stated as further work in many publications on Capon beamforming for medical ultrasound imaging the past decade. In \textbf{Paper I} arrays with no more than 32 elements were investigated. A linear array for cardiac ultrasound imaging typically has 64 or more elements. In \textbf{Paper VI}, which is summarized in \textbf{Paper II}, it is shown that our implementation from  \textbf{Paper I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging. The matrices that have to be inverted become to large, and the frames that need to be processed per second are too many.

@@

\subsection{Paper IV}
\textbf{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}\\
\textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben\\
{\it Proc. SPIE Medical Imaging 2012, \todo{add pages}.}\\\\
The fourth paper propose an adaptive volume rendering method based on statistics derived from the blood pool of the left ventricle. Noise located in the blood pool tends to occlude cardiac tissue when rendered in 3D, and is often impossible to remove by adjusting global rendering parameters. This because ultrasound signals have high variability within blood as well as high variability within tissue, even when heavily smoothed. Delineation of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time. The final visualizations are compared, with respect to location of tissue, by using a state-of-the-art endocardium segmentation tool. The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, leading to a maximized amount of visible endocardium tissue. The proposed method is implemented on a GPU with real-time frame rates as the result. This results shows that with modern GPUs it is possible to add more advanced visualization to an ultrasound imaging system and still have real-time performance.

\subsection{Paper V}
\textbf{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}\\
\textbf{J.\:P.\:\AA{}sen} and S.\:Holm\\
{\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}.}\\\\
The fifth paper presents an interactive simulations tool capable of simulating dense pressure fields from ultrasound arrays in real time. Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing. However, dense simulations involve extensive calculation and computational time. With the presented simulator we aim at reducing the time needed for calculating pressure fields by means of GPU acceleration. The tool is based on Huygens' principle which describes diffraction caused by a slit as the superposition of several point sources located inside the opening. Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points. How the source and observation points are laid out is left for the user to decide. Point sources positioned along a line will for instance simulate an ultrasound array. The simulator is linked with both a Paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations. The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox \todo{add citation}} version. Second, the paint-like interface provides a neat way of demonstrating array beamforming principle in real time. We believe this to be of great value both for array designers and teachers.

@@

In \textbf{Paper II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU. However, we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.  The paper does not provide any detailed evaluation of the method's \textit{in vivo} performance. Yet, initially it has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images. The improvements seen are minimal, especially with the number of calculations in mind. The beamformers are clearly able to decrees the lateral width of lateral point scatterers and thin structures \textit{in vivo}, and to sharpen step edges. On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper II}). Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs. The oversampling approach presented in \textbf{Paper III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small. Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles. In future work a detailed investigation of why some results, especially the contrast, obtained in simulations are not transferable to \textit{in vivo} needs to be performed. It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper II}. Both to justify all our computations, and to show an image which has both highly improved resolution and contrast. As pointed out at the end of \textbf{Paper II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming. Cardiac ultrasound imaging was selected because it is the medical ultrasound modality which requires most calculations per second. An application where the focus is detection of closely spaced point targets, might be better suited. A better suited adaptive beamformer for cardiac applications might be the low-complexity adaptive beamformer (\nom{LCA beamformer}{Low complexity adaptive beamformer}), where the Capon optimization problem is applied on a set of predefined windows. This method has linear complexity and shares many properties with the Capon beamformer. If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less computationally costly. Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally.

>>>>

\subsection{Paper II}
\textbf{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}\\
\textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C\:Nilsen, A.\:Austeng and S.\:Holm\\
{\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. 76-85, Jan. 2014.}\\\\
The second paper aims at implementing real-time Capon beamforming for cardiac ultrasound imaging. Achieving this will facilitate further study of the method's \textit{in vivo} performance. This is something which has been stated as further work in many publications on Capon beamforming for medical ultrasound imaging the past decade. In \textbf{Paper I} arrays with no more than 32 elements were investigated. A linear array for cardiac ultrasound imaging typically has 64 or more elements. In \textbf{Paper VI}, which is summarized in \textbf{Paper II}, it is shown that our implementation from  \textbf{Paper I} does not reach the level of throughput required for real-time Capon beamforming of cardiac ultrasound imaging. The matrices that have to be inverted become to large, and the frames that need to be processed per second are too many.

@@

\subsection{Paper IV}
\textbf{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}\\
\textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben\\
{\it Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008.}\\\\
The fourth paper propose an adaptive volume rendering method based on statistics derived from the blood pool of the left ventricle. Noise located in the blood pool tends to occlude cardiac tissue when rendered in 3D, and is often impossible to remove by adjusting global rendering parameters. This because ultrasound signals have high variability within blood as well as high variability within tissue, even when heavily smoothed. Delineation of the blood pool is done by a state estimation algorithm, capable of tracking the endocardium in real time. The final visualizations are compared, with respect to location of tissue, by using a state-of-the-art endocardium segmentation tool. The paper presents both quantitative and qualitative results supporting the method's improved capabilities of reducing tissue dropouts and spurious structures inside the blood pool, leading to a maximized amount of visible endocardium tissue. The proposed method is implemented on a GPU with real-time frame rates as the result. This results shows that with modern GPUs it is possible to add more advanced visualization to an ultrasound imaging system and still have real-time performance.

\subsection{Paper V}
\textbf{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}\\
\textbf{J.\:P.\:\AA{}sen} and S.\:Holm\\
{\it Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646.}\\\\
The fifth paper presents an interactive simulations tool capable of simulating dense pressure fields from ultrasound arrays in real time. Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing. However, dense simulations involve extensive calculation and computational time. With the presented simulator we aim at reducing the time needed for calculating pressure fields by means of GPU acceleration. The tool is based on Huygens' principle which describes diffraction caused by a slit as the superposition of several point sources located inside the opening. Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points. How the source and observation points are laid out is left for the user to decide. Point sources positioned along a line will for instance simulate an ultrasound array. The simulator is linked with both a Paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations. The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox \todo{add citation}} version. Second, the paint-like interface provides a neat way of demonstrating array beamforming principle in real time. We believe this to be of great value both for array designers and teachers.

@@

In \textbf{Paper II} we focused on achieving real time Capon processing for cardiac ultrasound by implementing the BS-Capon beamformer on a GPU. However, we also present the first medical ultrasound loops processed with ES- and BS-Capon beamforming.  The paper does not provide any detailed evaluation of the method's \textit{in vivo} performance. Yet, initially it has been proven hard to transfer all results obtained in simulations to \textit{in vivo} images. The improvements seen are minimal, especially with the number of calculations in mind. The beamformers are clearly able to decrees the lateral width of lateral point scatterers and thin structures \textit{in vivo}, and to sharpen step edges. On the other hand, the overall contrast is visually not improved and sometimes it is worse (Fig.\,6b of \textbf{Paper II}). Since the Capon beamformer only affect directional noise, it is clear that contrast will get worse in low \nom{SNR}{Signal-to-noise ratio} situations when signal cancellation occurs. The oversampling approach presented in \textbf{Paper III} might improve on this, but it will only reduce signal cancellation if the phase shifts across the aperture are close to linear and small. Another issue is that bright points with a wide lateral profile exhibits better visual contrast when surrounded by speckle noise than points with narrow lateral profiles. In future work a detailed investigation of why some results, especially the contrast, obtained in simulations are not transferable to \textit{in vivo} needs to be performed. It will also be crucial to show larger improvements \textit{in vivio} than what is presented in \textbf{Paper II}. Both to justify all our computations, and to show an image which has both highly improved resolution and contrast. As pointed out at the end of \textbf{Paper II}, cardiac ultrasound imaging might not be the best modality for Capon beamforming. Cardiac ultrasound imaging was selected because it is the medical ultrasound modality which requires most calculations per second. An application where the focus is detection of closely spaced point targets, might be better suited. A better suited adaptive beamformer for cardiac applications might be the low-complexity adaptive beamformer (\nom{LCA}{Low complexity adaptive (beamformer)} beamformer), where the Capon optimization problem is applied on a set of predefined windows. This method has linear complexity and shares many properties with the Capon beamformer. If only minor improvements are obtained, it would be easier to ignore this fact if the algorithm is less computationally costly. Finally, it would also be interesting to see how the Eigenspace-based Capon beamformer performs on cardiac images, and how its absence of a distortionless criterion will influence the image spatially and temporally.
